---
title: "Correlation Models"
output: html_document
---

You can do this on your own data, or on included data here.
```{r}
source("~/Desktop/Phylomethods/PhyloPlasticitySals/R/packages.R")  # loads packages

tree.primates <- read.tree(text="((((Homo:0.21,Pongo:0.21):0.28,Macaca:0.49):0.13,Ateles:0.62):0.38,Galago:1.00);") #using examples from ape ?pic


X <- c(4.09434, 3.61092, 2.37024, 2.02815, -1.46968)
Y <- c(4.74493, 3.33220, 3.36730, 2.89037, 2.30259)
names(X) <- names(Y) <- c("Homo", "Pongo", "Macaca", "Ateles", "Galago")
pic.X <- pic(X, tree.primates)
pic.Y <- pic(Y, tree.primates)
```

Now, positivitize the contrasts and do a regression through the origin.
```{r}
data(primates)
ls()
print(primates)
require(phytools)
```

Just to make this a better dataset for our exercise, I’m going to change some of the states (I want to have all four trait combinations present). For actual analyses, of course, DO NOT INVENT YOUR DATA.

First, a review of discrete state models:
```{r}
primates$trait[which(grepl("Hylobates",primates$trait[,1])),2]<-1

trait1<-primates$trait[,2]
names(trait1)<-primates$trait[,1]
primates$tree <- ape::multi2di(primates$tree)
plotSimmap(make.simmap(primates$tree, trait1), pts=FALSE, fsize=0.8)
rate.mat.er<-corHMM:::rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=2, model="ER")
print(rate.mat.er)

```

What does this matrix mean?
```{r}
pp.er<-corHMM(primates$tree,primates$trait[,c(1,2)],rate.cat=1,rate.mat=rate.mat.er,node.states="marginal")
print(pp.er)

#its an equal transition rate with one traits and two states. The summation of the rates are optimized 
```

What do these results mean?
```{r}
rate.mat.ard<-corHMM:::rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=2, model="ARD") #all rates differ
#rate.mat.er<-corHMM:::rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=2, model="ER") #all rates are equal


print(rate.mat.ard)
```

And these?
```{r}
pp.ard<-corHMM(primates$tree,primates$trait[,c(1,2)],rate.cat=1,rate.mat=rate.mat.ard,node.states="marginal")
#pp.er<-corHMM(primates$tree,primates$trait[,c(1,2)],rate.cat=1,rate.mat=rate.mat.er,node.states="marginal")
print(pp.ard)
```

which model is better?

#The ER, all rates are the same, model is better because it has a lower AIC

Now let’s look at multiple traits.

This is a matrix with four states
```{r}
rate.mat.er.4state<-corHMM:::rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ER")
print(rate.mat.er.4state)
```

Convert the two binary traits into a single four character state
```{r}
fourstate.trait<-rep(NA,Ntip(primates$tree))
for(i in sequence(Ntip(primates$tree))) {
    if(primates$trait[i,2]==0 && primates$trait[i,3]==0) {
        fourstate.trait[i]<-0
    }
    if(primates$trait[i,2]==0 && primates$trait[i,3]==1) {
        fourstate.trait[i]<-1
    }
    if(primates$trait[i,2]==1 && primates$trait[i,3]==0) {
        fourstate.trait[i]<-2
    }
    if(primates$trait[i,2]==1 && primates$trait[i,3]==1) {
        fourstate.trait[i]<-3
    }
}
fourstate.data<-data.frame(Genus_sp=primates$trait[,1], T1=fourstate.trait)

print(rayDISC(primates$tree, fourstate.data, ntraits=1, model="ER", node.states="marginal"))
print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat=rate.mat.er.4state, node.states="marginal", model="ARD"))
rate.mat.ard.4state<-corHMM:::rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ARD")
print(rate.mat.ard.4state)
```

Now let’s make the equivalent of a GTR matrix:
```{r}

#GTR = general time reversable 

rate.mat.gtr.4state<-rate.mat.ard.4state
rate.mat.gtr.4state<-corHMM:::rate.par.eq(rate.mat.gtr.4state, c(1,4))
rate.mat.gtr.4state<-corHMM:::rate.par.eq(rate.mat.gtr.4state, c(2,6))
rate.mat.gtr.4state<-corHMM:::rate.par.eq(rate.mat.gtr.4state, c(3,8))
rate.mat.gtr.4state<-corHMM:::rate.par.eq(rate.mat.gtr.4state, c(4,6))
rate.mat.gtr.4state<-corHMM:::rate.par.eq(rate.mat.gtr.4state, c(5,7))
rate.mat.gtr.4state<-corHMM:::rate.par.eq(rate.mat.gtr.4state, c(6,7))
print(rate.mat.gtr.4state)

print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat= rate.mat.gtr.4state, node.states="marginal", model="ARD"))  #same as corHMM but modified to use more than two states 
```

Now make a model like Pagel 1994
```{r}
#pagel model you can only go one direction 
print(corHMM:::rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=2, nstates=2, model="ARD"))
rate.mat.pag94<-corHMM:::rate.par.drop(rate.mat.ard.4state, drop.par=c(3,5,8,10))
print(rate.mat.pag94)
```

Now that you have some introduction, there are two routes:

##Route 1

Construct a model to test if state 1 can never be lost 
```{r}
rate.mat.minus.state.one <- corHMM:::rate.par.drop(rate.mat.ard, drop.par=1) #sets 2(1)- 1(0) transition to NA --- binary model from above 
rate.mat.minus.state.one.drop.NA <- rate.mat.minus.state.one
rate.mat.minus.state.one.drop.NA[is.na(rate.mat.minus.state.one.drop.NA)] <- 0
primates$trait2 <- primates$trait
primates$trait2[,2] <- primates$trait2[,2]+1

```

Experiment with the effects of frequencies at the root.
```{r}
maddfitz.ARD<-corHMM:::corHMM(primates$tree,primates$trait[,c(1,2)],rate.cat=1,rate.mat=rate.mat.ard, node.states="marginal", root.p="maddfitz")
yang.ARD <-corHMM:::corHMM(primates$tree,primates$trait[,c(1,2)],rate.cat=1,rate.mat=rate.mat.ard, node.states="marginal", root.p="yang")

maddfitz.ARD$AIC
yang.ARD$AIC

```



Create and use a model to see if transitions from 00 go to 11 only via 01.
```{r}
rate.mat.one.direction<-corHMM:::rate.par.drop(rate.mat.ard.4state, drop.par=c(7,10))

pp.every.rate<- rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat= rate.mat.ard.4state, node.states="marginal", model="ARD")
pp.every.rate.one.direction <- rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat= rate.mat.one.direction, node.states="marginal", model="ARD")

print(pp.every.rate$AIC) 
print(pp.every.rate.one.direction$AIC) 

#The model only allowing for a single path (pp.every.rate.one.direction) technically has a lower AIC, but the difference is negligible (i.e. less than or equal to four)
```

##Route 2

Maddison and FitzJohn (2015) pretty convincingly show (to me) that Pagel (1994) is just not a good method. Ok. So work on a fix. They point to Read and Nee (1995) as a low power but possible solution. Look at their appendix, especially, and write an implementation.