---
title: "Discrete Trees"
output: html_document
---
Body size in Salamanders. Extract and trim discrete data. 

```{r}
source("~/Desktop/Phylomethods/PhyloPlasticitySals/R/packages.R")  # loads packages
source("~/Desktop/Phylomethods/PhyloPlasticitySals/R/functions.R") # defines the create_plot() function
source("~/Desktop/Phylomethods/PhyloPlasticitySals/data/tree_and_data.R") # clean data 


pruned_tree <- PruneTree(tre,sal.data2)
avg.size <- male.female.avg.size(sal.data2)
size.class <- Size.Class.Female(avg.size)

all.discrete<- cbind(avg.size[,1], size.class)


```

Eyeball-check of the data 
```{r}

VisualizeData(phy=pruned_tree, data = all.discrete)

```


```{r}
#View(all.discrete)
levels(all.discrete) <- c('Large','Small')
cleaned.discrete.phyDat <- phangorn::phyDat(all.discrete, type="USER", levels = levels(all.discrete))
anc.p <- phangorn::ancestral.pars(pruned_tree, cleaned.discrete.phyDat)
plotAnc(pruned_tree, anc.p, 1)


##Do you see uncertianty? Yes some of the nodes contain split character traits (i.e. red/blue)

```

Likelihood estimates
```{r}
anc.ml <- ancestral.pml(pml(pruned_tree, cleaned.discrete.phyDat), type="ml")
summary(anc.ml)

plotAnc(pruned_tree, anc.ml, 1)
```

How does this differ from parsimony?
#Parsimony selects the simplest explanation (fewest number of state changes) to explain the observed data. 

Why does it differ from parsimony?
# The likelihood estimates rely more on character states of the tips (Large/Small) as opposed to the likelihood of each branch. 

What does uncertainty mean?
# There is uncertaininty regarding the scoring of characters at the tips and internal nodes. 

Now, to the biological questions. For many of these, corHMM will be a useful package. Do the following analyses:

How can you estimate transition rates between states? Do it.
How could you examine if transition rates are equal? 
Think about the Lewis (2001) MKV model. Are your traits all variable? Will using this make sense for your data? Try using it. Do results change?

```{r}

mod1<-fitDiscrete(pruned_tree, size.class, model = 'ER')
size_class2<-as.numeric(size_class)
is.rooted(pruned_tree)
tree<-unroot(pruned_tree)
is.rooted(tree)
mod2<-lewisMkv(tree, size_class2)

```


How could you test order of state evolution?

####How can you estimate transition rates between states ? By using the fitdiscrete function in geiger(## we can do that by fitting a  model with different backward & forward rates

```{r}

ard.mod<-fitDiscrete(pruned_tree, size.class, model = 'ARD')
print(ard.mod)

````

